---
title: "Learning cyclic causal graphs with backShift"
author: "Christina Heinze"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{backShift demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

status: process
---

Example for running [backShift](http://arxiv.org/abs/1506.02494), an algorithm 
to estimate the connectivity
matrix of a directed (possibly cyclic) graph with hidden variables. The
underlying system is required to be linear and we assume that observations
under different shift interventions are available.


## Options

### Preliminaries

```{r, message = FALSE}
# load package
require(backShift)
require(ggplot2)
require(fields)

# set seed
seed <- 1
set.seed(seed)
```

### Algorithm

backShift exploits differences between covariance or Gram matrices. To use 
covariance matrices set `useCov` to `TRUE`.

```{r}
# use covariance matrix instead of Gram matrix
useCov <- TRUE 
```

### Stability selection

backShift can be run with or without [stability selection](http://arxiv.org/abs/0809.2932). 
If stability selection should not be used, set `EV` to 0.

```{r}
# bound on expected number of false selections for stability selection
EV <- 2
# selection threshold for stability selection
thres <- 0.75 
```

To visualize the point estimate of the connectivity matrix, the coefficients are 
thresholded at the absolute value given by `thres.pe`. So edges with coefficients 
smaller than `thres.pe` in absolute value are not displayed.

```{r}
# threshold for point estimate 
thres.pe <- 0.25
```


### Simulation

Simulate a data set with `simulateInterventions()` and `generateA()` or use the 
connectivity matrix provided through `data("exampleAdjacencyMatrix")`.

#### Connectivity matrix

Set whether to provide the adjacency matrix A -- e.g. by loading an example 
with `data("exampleAdjacencyMatrix")` or by creating an adjacency matrix yourself --
or whether to generate the connectivity matrix A with `generateA()`.

NOTE: The entry A_ij contains the edge from node i to node j.

```{r, results = "hide"}
# number of variables
p <- 10

# set whether to provide or to generate the adjacency matrix A 
providedA <- TRUE

if(providedA){
  data("exampleAdjacencyMatrix")
  A <- exampleAdjacencyMatrix
  p <- 10
}else{
  # parameters to generate A
  
  # should A be cyclic?
  cyclic <- TRUE
  # expected number of neighbors per node
  expNumNeigh <- 0.1*p 
  # range for coefficients
  minCoef <- 0.3
  maxCoef <- 0.8
  
  ## Generate A -------
  cat("Generating A...\n") 
  A.gen.result <- generateA(p, expNumNeigh, minCoef, maxCoef, cyclic)
  A <- A.gen.result$A
  cat("A has a cycle of size", A.gen.result$sizeCycle, "\n") 
}
```

#### Observations under different shift interventions

```{r, results = "hide"}
# number of observations
n <- 10000
# number of environments
G <- 10 
# also simulate observational data?
simulateObs <- TRUE 
# should hidden vars be included?
hidden <- FALSE 
# should the location of the interventions be known?
knownInterventions <- FALSE 
# if the location of the interventions is known, how many vars. should
# be intervened on in each environment (as a fraction of p)
fracVarInt <- 0.5
# multiplier for interventions (m_I in manuscript)
intMult <- 1.5
# multiplier for interventions (m_e in manuscript)
noiseMult <- 1
# simulate non-Gaussian noise? 
nonGauss <- FALSE

## Simulate data -------
cat("Simulating the data...\n") 
simulation.res <- simulateInterventions(n, p, A, G, intMult, noiseMult, 
                                        nonGauss, hidden, knownInterventions, 
                                        fracVarInt, simulateObs, seed)

# extract X, environment vector and index of observational data
X <- simulation.res$X
env <- simulation.res$environment
baseInd <- simulation.res$configs$indexObservationalData
```

## Running backShift

```{r}
## Run backShift -------
backshift.res <- backShift(X, env, covariance=useCov, ev=EV, threshold=thres, 
                        baseSettingEnv = baseInd, tolerance = 1e-6, 
                        verbose = FALSE)
```


## Results

### Estimated graphs

```{r, fig.show='hold', fig.width=6, fig.height=6, fig.align='center', results = "hide"}
# extract estimates
Ahat <- backshift.res$Ahat
Ahat.structure <- backshift.res$AhatAdjacency

# compute performance metrics and plot result
cat("Plotting true graph... \n") 
plotGraphEdgeAttr(estimate = A, plotStabSelec = FALSE, labels = colnames(A), 
                  thres.point = 0, thres.stab = thres, main = "True graph")
```

```{r, fig.show='hold', fig.width=6, fig.height=6, fig.align='center', results="hide"}
cat("Plotting point estimate, thresholded at", thres.pe,"... \n") 
plotGraphEdgeAttr(estimate = Ahat, plotStabSelec = FALSE, labels = colnames(A), 
                  thres.point = thres.pe, thres.stab = thres, 
                  main = paste("Point estimate thresholded at", thres.pe))
```

```{r, fig.show='hold', fig.width=6, fig.height=6, fig.align='center', results = "hide"}
cat("Plotting stability selection result... \n") 
plotGraphEdgeAttr(estimate = Ahat.structure, plotStabSelec = TRUE, 
                  labels = colnames(A), thres.point = thres.pe, 
                  edgeWeights = Ahat, thres.stab = thres, 
                  main = "Stability selection result")
```

### Metrics

```{r, results = "asis"}
# metrics for point estimate, thresholded at thres.pe
metricsThresholdedA <- metricsThreshold(A, Ahat, thres = thres.pe)

# metrics for stability selection result
metricsStabSelection <- metricsThreshold(A, Ahat.structure, thres = 0)
```

```{r, results = "asis", echo = FALSE}
colnames(metricsThresholdedA)[3] <- colnames(metricsStabSelection)[3] <- "TPR/Recall"
knitr::kable(metricsThresholdedA, digits = 4, caption = paste("Metrics for point estimate, thresholded at", thres.pe))
knitr::kable(metricsStabSelection, digits = 4, caption = "Metrics for stability selection result")
```

## Estimating the intervention variances

The location and the strength of the shift interventions can be estimated from the
data (up to an offset). These are returned by the function `backShift()` as a 
($G \times p$)-dimensionalmatrix `varianceEnv` where $G$ is the number of environments and
$p$ is the number of variables. The ij-th entry contains the difference
between the estimated intervention variance of variable j in environment i
and the estimated intervention variance of variable j in the base setting
(given by input parameter `baseSettingEnv`).

```{r, message=FALSE, fig.width=7.5, fig.height=7.5}
plotInterventionVars(backshift.res$varianceEnv, simulation.res$interventionVar)
```


## Checking the model assumptions

We can check the model assumptions to some extent by the success or failure of the joint 
diagonalization procedure. The plots below show that the joint diagonalization succeeded
for all environments as all matrices are diagonal. The values are scaled to [0,1] 
to allow for comparability accross plots.

```{r, fig.width=3, fig.height=3}
for(i in 1:G){
  plotDiagonalization(estConnectivity = backshift.res$Ahat, X = X, env = env, whichEnv = i)
}
```
